\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{geometry}
\geometry{margin=1in}

\title{Absence of the Spin-Glass Phase in Dense Associative Memories}
\author{}
\date{}

\begin{document}
\maketitle

\section{Spin-Glass Phase in the Classical Hopfield Model}

In the classical Hopfield model~\cite{Amit1987}, the coupling matrix is
\begin{equation}
  J_{ij} = \frac{1}{N} \sum_{\mu=1}^{p} \xi_i^\mu \xi_j^\mu,
\end{equation}
and the local field acting on spin $S_i$ in a state aligned with pattern $\nu$ is
\begin{equation}
  h_i = \sum_{j \neq i} J_{ij} S_j
      = \xi_i^\nu (1 + \delta_i),
  \qquad
  \delta_i = \frac{1}{N} \sum_{j \neq i} \sum_{\mu \neq \nu}
             \xi_i^\mu \xi_j^\mu \xi_j^\nu,
\end{equation}
where $\delta_i$ is cross-talk noise with variance
$\langle \delta_i^2 \rangle = (p-1)/N$.
For $p = \alpha N$, this noise is $O(\sqrt{\alpha})$ and independent of $N$.

\subsection{Order parameters}

The mean-field theory uses two order parameters:
\begin{itemize}
  \item Overlap (magnetization):
    $m^\mu = \frac{1}{N} \sum_i \langle S_i \rangle \xi_i^\mu$
    --- measures alignment with pattern $\mu$.
  \item Edwards--Anderson parameter:
    $q = \langle N^{-1} \sum_i \langle S_i \rangle^2 \rangle$
    --- measures freezing (non-ergodicity).
\end{itemize}

\subsection{Phase diagram (Amit, Gutfreund, Sompolinsky 1987)}

The system exhibits three phases:
\begin{enumerate}
  \item \textbf{Paramagnetic (P):} $m = 0$, $q = 0$.
        Above $T_g = 1 + \sqrt{\alpha}$, the system is ergodic.

  \item \textbf{Spin-glass (SG):} $m = 0$, $q > 0$.
        Below $T_g$ (for $\alpha > \alpha_c \approx 0.138$ at $T = 0$),
        the system freezes into a random state with no macroscopic overlap
        with any stored pattern.
        The SG order parameter satisfies $q \simeq T_g - T$ near the transition.

  \item \textbf{Retrieval / Ferromagnetic (FM):} $m \neq 0$, $q > 0$.
        Below $T_M(\alpha)$, the system retrieves a stored pattern.
        For $\alpha < \alpha_c$, this is the ground state.
\end{enumerate}

The SG phase arises because the coupling matrix $J_{ij}$ is a rank-$p$
perturbation of what is effectively a random symmetric matrix.
From random matrix theory, $J$ has $p$ outlier eigenvalues (corresponding to
the stored patterns) sitting on a \emph{semicircular bulk} of width
$\sim 2\sqrt{\alpha}$. When $\alpha > \alpha_c$, the outliers merge into the bulk,
and the random-matrix structure dominates the energy landscape, creating an
exponentially large number of metastable states --- the spin-glass phase.

\section{Dense Associative Memories: LSE and LSR}

In dense (modern) associative memories, the energy function is qualitatively
different.  For the two models considered here:

\subsection{LSE (Log-Sum-Exp)}
\begin{equation}
  E_{\mathrm{LSE}}(\mathbf{S})
    = -\log \sum_{\mu=1}^{p}
      \exp\!\Bigl(\tfrac{1}{2}\,(\boldsymbol{\xi}^\mu \cdot \mathbf{S})^2 / N\Bigr).
\end{equation}

\subsection{LSR (Log-Sum with $b = 2 + \sqrt{2}$)}
\begin{equation}
  E_{\mathrm{LSR}}(\mathbf{S})
    = -\frac{1}{b}\,\log \sum_{\mu=1}^{p}
      \exp\!\Bigl(\tfrac{b}{2}\,(\boldsymbol{\xi}^\mu \cdot \mathbf{S})^2 / N\Bigr),
  \qquad b = 2 + \sqrt{2}.
\end{equation}

The $\exp(\cdot)$ inside the $\log\text{-sum}$ acts as a \textbf{soft-max}:
it exponentially amplifies the overlap with the best-matching pattern
and suppresses all others.

\section{Why No Spin-Glass Phase in Dense AM}

\subsection{Suppression of cross-talk}

In the Hopfield model, the local field is \emph{linear} in the overlaps:
\begin{equation}
  h_i^{\text{Hopf}} = \frac{1}{N}\sum_\mu m^\mu \xi_i^\mu
    = \frac{m^\nu}{N}\,\xi_i^\nu + \underbrace{\frac{1}{N}\sum_{\mu \neq \nu} m^\mu \xi_i^\mu}_{\text{cross-talk noise}}.
\end{equation}
All $p$ patterns contribute equally to the local field, regardless of their
overlap magnitude. The cross-talk noise scales as $\sqrt{\alpha}$ and generates
the frustrated landscape responsible for the SG phase.

In dense AM, the effective local field is dominated by the
\textbf{exponentially weighted} leading pattern:
\begin{equation}
  h_i^{\text{dense}} \propto
    \frac{\sum_\mu (\boldsymbol{\xi}^\mu \cdot \mathbf{S})\,\xi_i^\mu \,
    e^{\frac{b}{2}(\boldsymbol{\xi}^\mu \cdot \mathbf{S})^2/N}}
    {\sum_\mu e^{\frac{b}{2}(\boldsymbol{\xi}^\mu \cdot \mathbf{S})^2/N}}.
\end{equation}
The softmax weights $\propto e^{b\,N\,(m^\mu)^2/2}$ ensure that the pattern with
the largest overlap exponentially dominates the sum.
All other overlaps are suppressed by a factor
$\sim e^{-bN[(m^{\max})^2 - (m^\mu)^2]/2}$, which vanishes exponentially in $N$.

\subsection{No frustrated landscape at high $\alpha$}

The key consequence: when $\alpha$ exceeds the critical capacity $\alpha_c$,
the stored patterns become destabilized, but the cross-talk noise
\emph{does not create its own metastable states}.

\begin{itemize}
  \item \textbf{Hopfield:}
    $\alpha > \alpha_c$ $\Rightarrow$ patterns unstable, but the linear cross-talk
    creates $\sim e^{cN}$ random metastable states $\Rightarrow$ SG phase
    ($m = 0$, $q > 0$).

  \item \textbf{Dense AM:}
    $\alpha > \alpha_c$ $\Rightarrow$ patterns unstable, and the exponential
    suppression kills the diffuse noise that would form SG states.
    No metastable states survive $\Rightarrow$ paramagnetic phase
    ($m = 0$, $q = 0$).
\end{itemize}

Physically, the $\log$-$\sum$-$\exp$ energy landscape has deep, narrow wells
around stored patterns (exponentially deep in $N$), but the regions between
wells are \emph{smooth} --- no rugged random-matrix structure.
When the wells disappear at high $\alpha$, the landscape becomes flat, and
the system is ergodic.

\subsection{Direct R $\to$ P transition}

The phase diagram of dense AM therefore contains only two phases:
\begin{center}
\begin{tabular}{lcc}
  \hline
  Phase & $\varphi/\varphi_{\mathrm{th}}(T)$ & $q_{\mathrm{EA}}$ \\
  \hline
  Retrieval (R)      & $\approx 1$   & $> 0$ \\
  Paramagnetic (P)   & $\approx 0$   & $\approx 0$ \\
  \hline
\end{tabular}
\end{center}
with a direct R $\to$ P transition at $\alpha = \alpha_c(T)$,
and no intermediate SG phase.

This is confirmed by our basin stability simulations (v7), which show
$\text{SG} = 0$ for both LSE and LSR across the entire explored
$(\alpha, T)$ range.

\subsection{Summary}

The absence of the spin-glass phase in dense AM is a direct consequence
of the nonlinear (exponential) energy function:
\begin{enumerate}
  \item The softmax mechanism concentrates the energy on the dominant overlap,
        suppressing cross-talk.
  \item Without cross-talk, there is no Sherrington--Kirkpatrick-like
        frustrated coupling structure.
  \item Without frustration, no exponentially many metastable states form.
  \item Therefore, when retrieval fails ($\alpha > \alpha_c$), the system
        goes directly to the paramagnetic phase.
\end{enumerate}

This is one of the key advantages of dense associative memories over the
classical Hopfield model: a cleaner phase diagram with no spin-glass
contamination.

\begin{thebibliography}{9}
\bibitem{Amit1987}
  D.~J.~Amit, H.~Gutfreund, and H.~Sompolinsky,
  ``Statistical mechanics of neural networks near saturation,''
  \textit{Ann.\ Phys.} \textbf{173}, 30--67 (1987).
\end{thebibliography}

\end{document}
