\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{bm}
\usepackage{geometry}
\usepackage{hyperref}
\geometry{a4paper, margin=2.4cm}

\newcommand{\bxi}{\boldsymbol{\xi}}
\newcommand{\bx}{\bm{x}}
\newcommand{\Bnet}{\beta_{\mathrm{net}}}
\newcommand{\athLSR}{\alpha_{\mathrm{th}}^{\mathrm{LSR}}}
\newcommand{\acLSE}{\alpha_c^{\mathrm{LSE}}}

\title{\bfseries Internal note: is the LSR boundary-driven threshold
0.347 (NeurIPS) actually the correct analog of the LSE saddle-driven
threshold 0.6226?}
\author{}
\date{Drafted as a reference for referee responses to the AAAI submission.}

\begin{document}
\maketitle

\section*{Anticipated referee question}

The AAAI submission compares two thermodynamic-limit capacity numbers
under the exact spherical density:
\begin{itemize}
  \item LSE (this work): $\acLSE \approx 0.6226$, derived from a saddle of
        the spurious-sum integrand at the golden ratio
        $\varphi^{*} = (\sqrt{5}-1)/2$.
  \item LSR (NeurIPS companion~\cite{petrova2026metastable}):
        $\athLSR \approx 0.347$, derived from the support-entry condition
        $M \cdot \Pr[\varphi_{1\mu} > \varphi_c] \simeq 1$.
\end{itemize}
A careful referee may object that these are not the same kind of quantity.
The NeurIPS number is the threshold at which the \emph{first} spurious
pattern enters the LSR kernel support, computed from a single-closest /
extreme-value argument. The AAAI number is the threshold at which the
\emph{cumulative} spurious sum saturates the retrieval term, computed from
a saddle of the integrand. If for LSR the cumulative contribution from the
band of patterns near $\varphi_c$ overtakes the single closest pattern,
one might worry that $\athLSR = 0.347$ is the wrong number to compare
against $\acLSE = 0.6226$, and that the LSR cumulative-band saturation
threshold is different.

This note shows that, in the thermodynamic limit $N \to \infty$, the two
LSR criteria \emph{coincide}: NeurIPS's $0.347$ is simultaneously the
support-entry threshold and the boundary-driven cumulative-saturation
threshold, and it is therefore the correct apples-to-apples analog of the
LSE $0.6226$. The finite-$N$ correction is $O(\ln N / N)$ and explicitly
computed below.

\section{The two parallel derivations}

The spurious sum at the retrieval state $\bx = \bxi^1$ is
\begin{equation}
S \;=\; \sum_{\mu \neq 1} K\bigl(\varphi_{1\mu}\bigr)
   \;\approx\; M \int_{-1}^{1} f(\varphi)\,K(\varphi)\,\mathrm{d}\varphi,
\qquad f(\varphi) = C_N(1-\varphi^2)^{(N-3)/2},
\label{eq:S}
\end{equation}
where $K$ is the kernel and $C_N = 1/B((N-1)/2, 1/2)$. We compare with
the retrieval term $K_{\mathrm{ret}} = K(1)$ and call the breakdown the
condition $S \simeq K_{\mathrm{ret}}$.

\paragraph{LSE.}
Kernel $K_{\mathrm{LSE}}(\varphi) = \exp(\Bnet N \varphi)$,
retrieval value $K_{\mathrm{ret}} = \mathrm{e}^{\Bnet N}$. The integrand
exponent (per unit $N$) is
\begin{equation}
g_{\mathrm{LSE}}(\varphi) \;=\; \tfrac12 \ln(1-\varphi^2) \;+\; \Bnet\,\varphi.
\label{eq:gLSE}
\end{equation}
Stationarity at $\Bnet = 1$ gives $\varphi^{*2} + \varphi^{*} - 1 = 0$,
so $\varphi^{*} = (\sqrt{5}-1)/2$ and (using $1 - \varphi^{*2} = \varphi^{*}$)
\begin{equation}
g_{\max} \;=\; \tfrac12 \ln \varphi^{*} + \varphi^{*} \;\approx\; 0.3774.
\end{equation}
Setting $\alpha N + g_{\max} N = \Bnet N$ yields the saddle-driven LSE
threshold
\begin{equation}
\boxed{\;\acLSE \;=\; \Bnet - g_{\max} \;\approx\; 0.6226.\;}
\label{eq:acLSE}
\end{equation}

\paragraph{LSR.}
Kernel $K_{\mathrm{LSR}}(\varphi) = \max(0,\,1 - b + b\varphi) = b\,(\varphi-\varphi_c)\,\mathbf{1}_{\varphi \geq \varphi_c}$
with $\varphi_c = (b-1)/b$. Retrieval value $K_{\mathrm{ret}} = K_{\mathrm{LSR}}(1) = 1$.
The integrand on $[\varphi_c, 1]$ is $f(\varphi)\,b\,(\varphi-\varphi_c)$.
Taking logarithms, the exponent (per unit $N$) is
\begin{equation}
h_{\mathrm{LSR}}(\varphi)
\;=\; \underbrace{\tfrac12 \ln(1-\varphi^2)}_{\text{entropy, }O(1)}
   \;+\; \underbrace{\tfrac{1}{N}\ln\bigl[b(\varphi-\varphi_c)\bigr]}_{O(1/N),\,\text{subleading}}.
\label{eq:hLSR}
\end{equation}
To leading order in $N$ only the entropy contributes. The entropy
$\tfrac12 \ln(1-\varphi^2)$ is monotonically decreasing on $[\varphi_c, 1]$
(its global maximum at $\varphi = 0$ lies \emph{outside} the LSR support),
so the integrand-exponent maximum is at the left boundary $\varphi = \varphi_c$
with value
\begin{equation}
h_{\max} \;=\; \tfrac12 \ln(1-\varphi_c^2).
\end{equation}
Setting $\alpha N + h_{\max} N = 0$ yields the boundary-driven LSR threshold
\begin{equation}
\boxed{\;\athLSR \;=\; -h_{\max} \;=\; -\tfrac12 \ln(1-\varphi_c^2).\;}
\label{eq:athLSR}
\end{equation}
For $b = 2 + \sqrt{2}$ this is $\athLSR = \tfrac12 \ln 2 \approx 0.347$,
which is exactly the NeurIPS number.

\section{Why the numbers (0.6226 vs.\ 0.347) look so different}

Both formulas are instances of the same identity:
\begin{equation}
\boxed{\;\alpha_c \;=\; (\text{retrieval rate}) \;-\; (\text{integrand maximum value}).\;}
\end{equation}
The contrast in the numerical answers has two ingredients, not one.
\begin{enumerate}
  \item \textbf{Where the integrand maximum sits.}
        LSE's exponential kernel adds the linear term $\Bnet\varphi$ to
        the entropy, pushing the maximum inward to an interior saddle at
        the golden ratio $\varphi^{*}=0.618$. LSR's compactly-supported
        kernel contributes only an $O(1/N)$ correction; the leading-order
        maximum sits at the support boundary $\varphi_c = 0.707$ for
        $b = 2+\sqrt{2}$.
  \item \textbf{Retrieval-kernel scale.}
        LSE's $K_{\mathrm{ret}} = \mathrm{e}^{\Bnet N}$ supplies a rate
        $\Bnet$ on the right-hand side of the balance. LSR's
        $K_{\mathrm{ret}} = 1$ supplies a rate of $0$. Algebraically:
        $\acLSE = \Bnet - g_{\max}$ and $\athLSR = 0 - h_{\max}$.
\end{enumerate}

The values inserted into the boxed identity are:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lcccc}
\hline
                  & retrieval rate & integrand max position & integrand max value & threshold \\
\hline
LSE (this work)   & $\Bnet = 1$ & $\varphi^{*} = (\sqrt{5}-1)/2$ & $g_{\max} = +0.3774$ & $0.6226$ \\
LSR (NeurIPS)     & $0$         & $\varphi_c = (b-1)/b$          & $h_{\max} = -0.347$  & $0.347$ \\
\hline
\end{tabular}
\end{center}

The 0.6226-vs-0.347 contrast is therefore neither a saddle-vs-boundary
distinction alone nor a kernel-scale distinction alone: it is both.

\section{Cumulative band vs.\ single closest, the original worry}

The referee's intuition is correct that there is a band of patterns just
above $\varphi_c$ whose cumulative contribution exceeds that of the single
closest pattern. The point of this section is that the threshold extracted
from the cumulative-band calculation coincides with the threshold extracted
from the single-closest / support-entry calculation in the thermodynamic
limit.

\paragraph{Position of the cumulative band.}
The full LSR integrand is $f(\varphi)\,(\varphi-\varphi_c)$. Setting its
log-derivative to zero,
\begin{equation}
-\frac{(N-3)\varphi}{1-\varphi^2} + \frac{1}{\varphi-\varphi_c} = 0,
\end{equation}
and expanding around $\varphi_c$ with $\delta = \varphi-\varphi_c \ll 1$
gives the interior maximum at
\begin{equation}
\delta_{\star} \;\approx\; \frac{1-\varphi_c^{2}}{N\,\varphi_c} \;=\; O(1/N).
\end{equation}
So LSR \emph{does} have a formal interior maximum of the integrand, but its
position is glued to the support boundary in the $N \to \infty$ limit. The
Gaussian width of the saddle is $\sigma \sim 1/\sqrt{N}$.

\paragraph{Cumulative-band sum.}
Laplace integration around $\delta_{\star}$ gives
\begin{equation}
S_{\mathrm{LSR}} \;\sim\; M \, b \, f(\varphi_c) \cdot \frac{1}{N^{2}\varphi_c^{2}}
\;=\; \exp\!\bigl[(\alpha - \athLSR)\,N\bigr] \cdot \frac{b\sqrt{2}}{N^{3/2}\sqrt{\pi}\varphi_c}
\end{equation}
(using $f(\varphi_c) \sim \sqrt{N/(2\pi)} \, \mathrm{e}^{-\athLSR N}$
at leading order). Setting $S_{\mathrm{LSR}} = K_{\mathrm{ret}} = 1$:
\begin{equation}
\alpha_c^{\mathrm{cum,\,LSR}}(N)
\;=\; \athLSR \;+\; \frac{\tfrac{3}{2}\ln N + \tfrac12 \ln \tfrac{\pi}{2} + \ln(\varphi_c/b)}{N}
\;=\; \athLSR + O\!\left(\frac{\ln N}{N}\right).
\label{eq:athLSR-cum}
\end{equation}
For $N = 25$ and $b = 2+\sqrt{2}$ this gives a shift of about $+0.15$
relative to the leading-order value; for $N = 100$ it is about $+0.05$; in
the thermodynamic limit it vanishes.

\paragraph{Single-closest argument.}
NeurIPS works with the typical maximum overlap among $M$ patterns:
$\varphi_{\max}(\alpha) \approx \sqrt{1 - \mathrm{e}^{-2\alpha}}$ (from
inverting $M \int_{\varphi_{\max}}^{1} f \, \mathrm{d}\varphi = 1$). The
condition $\varphi_{\max} = \varphi_c$ gives precisely the leading-order
value of \eqref{eq:athLSR-cum}, with no $O(\ln N / N)$ correction. The
single-closest argument is the leading-order term of the cumulative one.

\paragraph{Why they agree at $N \to \infty$.}
Because the LSR integrand maximum sits \emph{at} the support boundary
(unlike the LSE case, where the saddle is in the interior at
$\varphi^{*} = 0.618$), the band of patterns and the single closest pattern
are not statistically separable structures: the band collapses onto the
boundary as $N \to \infty$, the width $\delta_{\star} \to 0$ as $1/N$, and
the number of patterns inside the band stays $O(1)$ at the threshold.
For LSE the saddle width is also $\sigma \sim 1/\sqrt{N}$ but its
\emph{position} is bounded away from any boundary (at $\varphi^{*} \approx 0.618$),
so the cumulative-band picture survives the thermodynamic limit and is
qualitatively distinct from any single-closest argument.

\section{Bottom line for the referee response}

There is no separate ``cumulative-band breakdown threshold for LSR''
distinct from NeurIPS's $\athLSR = 0.347$ in the thermodynamic limit.
$\athLSR = 0.347$ is simultaneously
\begin{enumerate}
  \item the support-entry threshold (first pattern enters the LSR kernel
        support), and
  \item the boundary-driven cumulative-saturation threshold (spurious sum
        first balances the retrieval term),
\end{enumerate}
to leading order in $N$. Its finite-$N$ correction is the $+ O(\ln N / N)$
shift in \eqref{eq:athLSR-cum}, which appears as a small additive offset
in the apparent boundary at the moderate $N$ accessible to honest Monte
Carlo and vanishes in the $N \to \infty$ limit. Comparing $0.6226$ (LSE,
this work) and $0.347$ (LSR, NeurIPS) is therefore an apples-to-apples
comparison of thermodynamic-limit, exact-density, first-noise-saturation
thresholds. The numerical contrast reflects both the interior-saddle vs.\
boundary-max distinction (the kernel-shape effect) and the
$\mathrm{e}^{\Bnet N}$ vs.\ $1$ scaling of the retrieval term (the
kernel-amplitude effect).

\begin{thebibliography}{9}
\bibitem{petrova2026metastable}
T.\ Petrova, E.\ Polyachenko, R.\ State.
\emph{Retrieval Metastability via Kramers Escape and Centroid Hopping}.
NeurIPS, 2026 (placeholder bib key).
\end{thebibliography}

\end{document}
